Applying State-free Reversible VAMPNets and Markov State Models to Learn Dynamics of DNA Oligonucleotides

### 2020-11-23 manuscript.tex
Updates from Overleaf

\usepackage[scaled=0.85]{beramono}
\usepackage[T1]{fontenc}
\usepackage{changepage}

@@

Despite rapid advances in DNA nanotechnology and a robust understanding of the associated thermodynamics, the sequence-dependent mechanisms of DNA hybridization are not fully understood. In this work, we investigate these dynamics by performing equilibrium coarse-grained simulations of oligonucleotide sequences with varied G:C placement. We employ State-Free Reversible VAMPnets to directly learn the slowest dynamical modes of each sequence and optimize Markov State Models (MSMs) construction. Furthermore, we perform elevated temperature simulations to recapitulate temperature-jump IR and FTIR data collected on the oligonucleotides. For repetitive sequences, we find a spectrum of slow dynamics associated with out-of-register base pairing and kinetically relevant transitions between these states. In contrast, G:C pairs near the center of the duplex induce more rapid fraying dynamics. In both cases, hybridization/dissociation mechanisms deviate from an “all-or-nothing” model. Our computational predictions are in excellent accord with experiment, and provide new fundamental understanding of the sequence-dependent kinetics and mechanisms of DNA hybridization.

@@

Over the last couple decades, DNA has proved to be much more than a vessel for genetic information. From sensing, to computing, to directed self-assembly, the programmable and predictable nature of DNA has unlocked numerous unforeseen nanotechnology applications \citep{Seeman2017DNANanotechnology, Adleman1994MolecularProblems, Rothemund2006FoldingPatterns, Gu2010ALine}. DNA dynamical behavior has become increasingly important in these technologies and a point of focus in fundamental biological processes such as transcription and gene regulation.  \citep{Deluca2020DynamicDevices, Cordes2010SensingTransfer, Naimark2020DNADevelopment}. Our understanding of DNA hybridization -- the formation of a duplex from two amenable single strands -- has been integral to these developments. The effect of sequence on hybridization thermodynamics has been rigorously explored, and nearest-neighbor calculations can account for many mismatched pairs, dangling ends, and other non-native bonding effects \citep{SantaLucia1998AThermodynamics, Santalucia2004TM}. Secondary DNA structures such as hairpins and G-quadruplexes have also been studied in depth and leveraged for nanotechnology applications \citep{Tsukanov2013DetailedOrigami, Mosayebi2014TheFormation, Mergny2019DNANanotechnology}. Although many experimental and computational studies have investigated DNA dynamical phenomena from the picosecond to millisecond range, the sequence-dependent mechanisms of hybridization and dissociation dynamics are not fully understood \citep{Yin2011KineticsHybridization, Xiao2019, Hinckley2014Coarse-grainedEffects, Sanstead2016, Porschke1971CooperativeTransition, Porschke1973ThermodynamicsPairs, Chen2007InfluenceHybridization, Craig1971ElaxationOligon}. Moreover, it is unclear the extent to which these processes evolve in an "all-or-nothing" fashion or if an ensemble metastable states facilitates the transition \citep{Araque2016LatticeCooperativity, Sikora2013ModelingIntermediates}.

Our understanding of hybridization dynamics has been built from decades of experiments -- such as temperature-jump, salt-jump, pH-jump, or other perturbative methods -- that drive DNA out of equilibrium and monitor relaxation processes in one direction \citep{Morrison1993SensitiveSolution, Wetmur1968KineticsDNA, Craig1971ElaxationOligon, Porschke1973ThermodynamicsPairs, Williams1989LaserDGCATGC, Narayanan2012ExploringMixing, Chen2007InfluenceHybridization, Sanstead2018DirectDehybridization}. More recently, single molecule diffusion and tethered multifluorophore assays have facilitated equilibrium analysis, however these results can be hampered by slow data collection rates and fluorescent tags altering strand dynamics, particularly for shorter oligomers \citep{Liu20173DSolution,  Schickinger2018TetheredHelices, Chen2008Base-by-baseSpectroscopy, Dupuis2013Single-moleculeHelices, Morrison1993SensitiveSolution}. Given that dynamic insights from these experiments are limited, several coarse-grained molecular dynamics (MD) models have been employed to gain further detail \citep{Romano2013DNADependence, Hinckley2013AnHybridization, Maciejczyk2014DNAModel, Markegard2015}. Although these models provide experimentally verified speed-ups compared to all-atom simulations, the long timescales on which DNA hybridization and dissociation events occur can still make these processes difficult to sample via direct simulation techniques \citep{Phys2014}. Instead, many previous studies of DNA hybridization have employed accelerated sampling methods such as metadynamics, umbrella sampling,  transition path sampling, and forward flux sampling \citep{Schmitt2013ExploringSurface, Sambriski2009,  Hoefert2011MolecularOligonucleotides, Romano2013DNADependence}. Simplified lattice models can expedite computation time and yield useful thermodynamics and transition state information, however some configurations may not be accessible and kinetics are less informative \citep{Araque2016LatticeCooperativity, Phys2019}. Other computational works use dramatically elevated temperature or denaturing solvent concentrations to induce one-way dissociation events \citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding}. Taken together, most experimental and computational work have studied certain aspects the overall dynamics process in one direction or in a biased fashion.

Recent studies have coupled experimental techniques with machine learning (ML) and  MD simulations to investigate and predict sequence-dependent kinetics \citep{Schickinger2018TetheredHelices, Zhang2018PredictingSequence}. Where these studies focus on association and dissociation kinetics alone, we seek to broaden our analysis into higher order dynamical processes and metastable intermediates. For example, metastable out-of-register or "shifted" base pairing in repetitive sequences have been documented in previous computational studies \citep{Phys2014, , Maciejczyk2014DNAModel, Araque2016LatticeCooperativity, Xiao2019}). The extent to which these states are kinetically relevant is difficult to determine experimentally, and appears highly sequences dependent. Frayed structures and dynamics have also been investigated in numerous computational and experimental studies \citep{Zgarbova2014BaseRNA, Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix}. Sanstead et al. highlighted the role of these dynamics during duplex dissociation, where the stability and relaxation timescale of frayed states was dictated by G:C base placement 10-mer oligonucleotide sequences \citep{Sanstead2016}. Together, these dynamics represent the most substantial deviations from all-or-nothing behavior that can be variably expressed via subtle sequence differences in short oligomers.

In this work, we study the same four sequences explored by Sanstead et al in an effort to uncover the sequence-dependent dynamics and their relation to metastable structures mentioned above \citep{Sanstead2016}. We use the coarse-grained 3 Sites per Nucleotide (3spn2) model to simulate hybridization and dissociation behavior near each sequence's melting temperature \citep{Hinckley2013AnHybridization}. We perform these analyses without biasing simulations or assuming that one processes is a strictly reversible version of the another. Furthermore, we leverage the properties of Markov State Models (MSMs) -- namely that conditional probability depends only on the current state of the system \citep{Pande2010EverythingAsk} -- to combine many independent and unbiased trajectories and develop an understanding of sequence-specific kinetics and thermodynamics. MSMs have recently been implemented to study mechanisms and microstate distributions of DNA hybridization  \citep{Jin2019, Xiao2019}, but the slowest sequence-dependent kinetics were not the focus of these studies. \citet{Pinamonti2017} used MSMs to compare the slowest dynamics of short RNA nucleotides and found that stacking timescales are highly sequence dependent \citep{Pinamonti2017}. We take a similar approach to study 10-mer DNA oligonucleotides and introduce State Free Reversible Vampnets (SRVs) to directly learn the slowest sequence-dependent dynamical modes \citep{Chen}. Furthermore, we integrate SRVs into the MSM pipeline by generating an optimized low dimensional basis in which microstate clustering can be performed. We show that SRV coordinates can be useful for both directly interpreting dynamical trends and for improving overall SRV-MSM quality when compared to more conventional methods such as time-structure independent components analysis (tICA).

We find that G:C base pair placement in decamer oligonucleotides has a substantial effect on dynamical behavior. By evaluating equilibrium trajectories we can study the relevance of metastable states during both the hybridization and dissociation process. Because SRVs generate an optimized low dimensional basis, we show that we can access higher resolution MSMs (about a 50\%reduction in required lag time) and generate more detailed models. Additionally, we can compare slow dynamical modes and metastable states between sequence-specific SRV-MSMs. Within these metastable dynamical states, we leverage diffusion maps to analyze the diversity of structures whose inter-conversion rate are too fast to produce unique slow modes. Finally, we run higher temperatures simulations to investigate the temperature-dependent nature of some experimentally relevant dynamical responses. Taken together, our analysis reflects similar results to previous computational and experimental DNA work, while elucidating new insights into sequence-dependent dynamics, metastable structures, and relative timescales.

@@

The 3 site per nucleotide (3spn2) coarse grained model was designed to accelerate DNA computation relative to all-atom models while maintaining experimental melting temperatures, stacking energies, and persistence lengths\citep{Hinckley2013AnHybridization}. Interactions sites are located at phosphate, sugar, and base centers of mass; anisotropic potentials are designed to model non-bonded interactions such as intra-strand base-stacking, inter-strand cross-stacking, and base pairing. The model has been validated against experimentally determined structural properties and hybridization rates, although no dynamic information was used to parameterize the model other than Langevin friction coefficients \citep{Hinckley2013AnHybridization, Hinckley2014Coarse-grainedEffects}. 3spn2 has widely-adopted to study numerous phenomenon including DNA packing in viral capsids, protein-DNA binding, and nucleosome unwrapping. \citep{Cordoba2017AIons, Lu2020OpenAWSEMSummary, Lequieu2016Tension-dependentUnwrapping}.

We initialized four sequences previously the subject of ultra-fast T-jump experimentation -- 5'ATATATATAT3' (AT-all), 5'GATATATATC3' (GC-end), 5'ATATGCATAT3' (GC-core), and 5'ATGATATCAT3' (GC-mix) -- along with their complementary strands in a periodic box of dimensions 7.774 nm \citep{Sanstead2016, Phys2014}. We specified a 240 mM implicit salt concentration and used a A Debye-Huckel approximation for electrostatic interactions within a 5 nm cutoff. We ran our simulations in the NVT ensemble and fixed temperature via a Langevin thermostat in order to model implicit solvent interactions \citep{Schneider1978Molecular-dynamicsTransitions}. Simulations were conducted at empirically determined  melting temperatures in order to maximize transitions between dissociated and hybridized states. We use a 20 fs time step and $13\times10^{9}$ steps in each simulation, saving frames every 100 ps. For each sequences, 40 simulations were performed in parallel, consuming about 24 serial CPU-hours on 28xIntel E5-2680v4 processors for each simulations. Half of runs were initialized from the hybridized state and half from a dissociated state. The first 10000 frames (1 $\mu$s of simulation time) of each simulation were removed, resulting in 40 x 250000 frames and a total of 1 ms simulation time per sequence. Depending on sequence, we observed between 40-100 hybridizing and dissociating transitions across all runs.

@@

All intermolecular pairwise distances for both oligonucleotides were calculated at each frame using the MDtraj software package \citep{McGibbon2015MDTraj:Trajectories}. Intermolecular distances removed translational and rotational symmetries, but additional symmetries arise from the self-complementary nature of these sequences -- both the sense and anti-sense strands are identical to each other. To account for this, we averaged permutable distances (45 pairs in total) together. This permutation reduction follows a similar procedure used in tICAgg coordinates construction \citep{Sengupta2019AutomatedSelf-assembly}. The VAMP-2 scoring method was employed to evaluate the kinetic variance of the feature set and optimize hyperparameters \citep{WuVariationalData, Mardt2018VAMPnetsKinetics}. The VAMP-2 score uses the covariances of a set of inputs estimates the transfer operator of a dynamical system, providing a robust and object means to evaluate various parameters and models.

@@

We compared scores for the intermolecular, intra+intermolecular,  and symmetrized intermolecular distances and found small differences between the three that varied by sequences, but overall we did not observe a loss in kinetic information when using the symmetrized feature set (\ref{fig:allseq_features_vamp2}). We found a substantial increase in VAMP-2 score when using reciprocal pairwise distances and chose reciprocal symmetrized coordinates to train the model. The smaller feature set enabled faster training times and better statistics over permutable distances, without appearing to suffer a loss in generality or model resolution. These features were normalized and passed into sequence-specific SRVs.

@@

SRVs were first developed by Chen et al. as a means to directly learn slow eigenfunctions of the transfer operator \citep{Chen}. The method is a descendant of VAMPnets \citep{Mardt2018VAMPnetsKinetics}, deep canonical correlation analysis \citep{Andrew2013DeepAnalysis},and extended dynamic mode decomposition with dictionary learning (EDMD-DL) \citep{Li2017ExtendedOperator}. The framework uses a twin-lobed artificial neural network to learn an optimal basis set for the variational approach to conformational dynamics (VAC) from which the leading eigenfunctions of the transfer operator are then estimated \citep{Noe2013ASystems}. The resulting orthogonal modes are associated with the slowest dynamical processes in a system, and can be used to interpret kinetic information directly (such as physical correlations and timescales) and to construct MSMs \citep{Sidky, Pande2010EverythingAsk}. SRVs provide robust nonlinear approximation and computation time that scales linearly with the amount of input data and are more powerful and efficient that tICA and ktICA approaches. This is a key attribute to our system as 10 million frames with 55 features in each frame are used for each sequence. The SRV framework has been tested on toy systems where the true eigenfunctions of the transfer operator are known and on small protein simulation data such as the WW-domain and Trp-cage mini-protein \citep{Chen, Sidky}. For these latter system, SRV-MSMs were contructed in order to find the stability of metastables states as well as transition probabilities between those states.

% might want to include these training params in the SI?

@@

MSMs are a powerful tool for interpreting large amounts of simulation data in a statistically robust and experimentally comparable way \citep{Phys2011MarkovValidation, Husic2018MarkovScience}. The technique relies on the discretitization of kinetically similar conformations into microstates and finds the conditional probability between states within some lag time. The reliance on conditional probabilities allows for many independent simulations (longer than the lag time) to be collectively interpreted. To take full advantage of the MSM frameworks, however, the input basis should be as kinetically meaningful as possible \citep{Pande2010EverythingAsk}. Because SRV eigenfunctions translate simulation features into their slowest kinetic representations, they are optimally suited as an MSM basis. To build our SRV-MSM framework, we employed the PyEmma MSM pipeline and generated independent models for each sequence \citep{Scherer2015PyEMMAModels}. In a similar approach to Sidky et al. we performed k-means microstate clustering, Bayesian MSM construction, and PCCA+ hierarchical macrostate assignments \citep{Sidky}. The number of microstates were determined by VAMP-2 score, and the SRV-MSM lag time was selected based on implied timescales convergence.  The number of PCCA+ macrostates was determined based on the characteristic of each system and will be discussed more in depth in the results. Additional step-by-step details on SRV-MSM construction for each sequence are provided in the supplemental information and on Github (https://github.com/mrjoness/...).

@@

% A Debye-Huckle screening potential was used to monitor account for ion concentration .

@@

\subsection{All SRV-MSMs}

@@

\subsubsection{SRV optimization and analysis}
In our analysis, we found that the AT-all sequence, given its repetitive structure and lack of GC-content, produced the simplest dynamics and displayed a clear spectral gap between modes. For this reason, we use this sequence a case study to work through our SRV-MSM pipeline step-by-step. Our first task was to identify the SRV lag time that was longer than the intrinsic Markov timescales of the system, yet short enough to resolve the dynamics of interest \citep{Phys2018MarkovValidation}. We found that most implied timescales converge at an SRV lag time of 1.2 ns. Next we selected an optimal number of SRV components to include in our analysis. After a certain point, higher order dynamical modes provide diminishing contributions the overall kinetic variance as measured by the Vamp-2 score, and the model can begin fitting on statistical noise in the trajectory data instead of the true dynamics \citep{McGibbon2015VariationalKinetics}. It is also more difficult to perform kmeans clustering on a high dimensional space, especially when those higher dimensions are less kinetically relevant \citep{Pande2010EverythingAsk}. For these reasons, the number of slow SRV components should be carefully selected based on the specific system of interest. As shown in figure \ref{fig:allseq_srv_crossval}), we see diminishing returns in the VAMP-2 score after five slow modes and select these modes as our optimized SRV basis.

>>>>

\usepackage[scaled=0.85]{beramono}
\usepackage[T1]{fontenc}
\usepackage{changepage}
\usepackage{mathrsfs}

@@

Despite rapid advances in DNA nanotechnology and a robust understanding of the associated thermodynamics, the sequence-dependent mechanisms of DNA hybridization are not fully understood. In this work, we investigate these dynamics by performing equilibrium coarse-grained simulations of oligonucleotide sequences with varied G:C placement. We employ State-Free Reversible VAMPnets to directly learn the slowest dynamical modes of each sequence and optimize Markov State Model (MSM) construction. Furthermore, we perform elevated temperature simulations to recapitulate temperature-jump IR and FTIR data collected on the oligonucleotides. For repetitive sequences, we find a spectrum of slow dynamics associated with out-of-register base pairing and kinetically relevant transitions between these states. In contrast, G:C pairs near the center of the duplex induce more pronounced fraying dynamics through which hybridization and dissociation are facilitated. In both cases, these mechanisms deviate from an “all-or-nothing” hybridization model. Our computational predictions show agreement with experiments, and provide new fundamental understanding of the sequence-dependent kinetics and mechanisms of DNA hybridization.

@@

Over the last couple decades, DNA has proved to be much more than a vessel for genetic information. From sensing, to computing, to directed self-assembly, the programmable and predictable nature of DNA has unlocked numerous unforeseen nanotechnology applications \citep{Seeman2017DNANanotechnology, Adleman1994MolecularProblems, Rothemund2006FoldingPatterns, Gu2010ALine}. In particular, DNA dynamical behavior has become increasingly important in developing these technologies and understanding fundamental biological processes such as transcription and gene regulation \citep{Deluca2020DynamicDevices, Cordes2010SensingTransfer, Naimark2020DNADevelopment}. A thorough understanding of the hybridization process -- the formation of a DNA duplex from two single strands -- and its associated dynamics is integral to further advancements in these fields. The effect of sequence on hybridization thermodynamics has been rigorously explored, and nearest-neighbor calculations can account for mismatched pairs, dangling ends, and other non-native bonding effects \citep{SantaLucia1998AThermodynamics, Santalucia2004TM}. Secondary DNA structures such as hairpins and G-quadruplexes have also been studied in depth and leveraged for nanotechnology applications \citep{Tsukanov2013DetailedOrigami, Mosayebi2014TheFormation, Mergny2019DNANanotechnology}. Although many experimental and computational studies have investigated DNA dynamical phenomena from the picosecond to millisecond range, the sequence-dependent mechanisms of hybridization and dissociation dynamics are not fully understood \citep{Yin2011KineticsHybridization, Xiao2019, Hinckley2014Coarse-grainedEffects, Sanstead2016, Porschke1971CooperativeTransition, Porschke1973ThermodynamicsPairs, Chen2007InfluenceHybridization, Craig1971ElaxationOligon}. Moreover, it is unclear the extent to which these processes evolve in an "all-or-nothing" fashion or if an ensemble metastable states facilitates the transition \citep{Araque2016LatticeCooperativity, Sikora2013ModelingIntermediates}.

Our understanding of hybridization dynamics has been built from decades of experiments -- such as temperature-jump, salt-jump, pH-jump, and other perturbative methods -- that drive DNA out of equilibrium and monitor relaxation processes in one direction \citep{Morrison1993SensitiveSolution, Wetmur1968KineticsDNA, Craig1971ElaxationOligon, Porschke1973ThermodynamicsPairs, Williams1989LaserDGCATGC, Narayanan2012ExploringMixing, Chen2007InfluenceHybridization, Sanstead2018DirectDehybridization}. More recently, single molecule diffusion and tethered multifluorophore assays have facilitated equilibrium analysis, however these results can be hampered by slow data collection rates and fluorescent tags effects on strand dynamics, particularly for shorter oligomers \citep{Liu20173DSolution,  Schickinger2018TetheredHelices, Chen2008Base-by-baseSpectroscopy, Dupuis2013Single-moleculeHelices, Morrison1993SensitiveSolution}. Given that dynamic insights from these experiments are limited, several coarse-grained molecular dynamics (MD) models have been employed to gain further detail \citep{Romano2013DNADependence, Hinckley2013AnHybridization, Maciejczyk2014DNAModel, Markegard2015}. Although these models provide experimentally verified speed-ups compared to all-atom simulations, the long timescales on which DNA hybridization and dissociation events occur can still make these processes difficult to sample via direct simulation techniques \citep{Phys2014}. Instead, many previous studies of DNA hybridization have employed accelerated sampling methods such as metadynamics, umbrella sampling,  transition path sampling, and forward flux sampling \citep{Schmitt2013ExploringSurface, Sambriski2009,  Hoefert2011MolecularOligonucleotides, Romano2013DNADependence}. Simplified lattice models can expedite computation time and yield useful thermodynamics and transition state information, however some configurations may not be accessible and kinetics are less informative \citep{Araque2016LatticeCooperativity, Phys2019}. Other computational works use dramatically elevated temperature or denaturing solvent concentrations to induce one-way dissociation events \citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding}. Taken together, most experimental and computational work have studied certain aspects the overall dynamics process in one direction or in a biased fashion.

Recent studies have coupled experimental techniques with machine learning (ML) and  MD simulations to investigate and predict sequence-dependent kinetics \citep{Schickinger2018TetheredHelices, Zhang2018PredictingSequence}. Where these studies focus on association and dissociation kinetics alone, we seek to broaden our analysis into higher order dynamical processes and metastable intermediates. For example, stable out-of-register or "shifted" base pairing in repetitive sequences have been documented in previous computational studies \citep{Phys2014, Maciejczyk2014DNAModel, Araque2016LatticeCooperativity, Xiao2019}. The extent to which these states are kinetically relevant is difficult to determine experimentally, and appears highly sequence-dependent. Frayed structures and dynamics have also been investigated in numerous computational and experimental studies \citep{Zgarbova2014BaseRNA, Nonin1995TerminalFraying, Nikolova2012ProbingSimulations, Andreatta2006UltrafastHelix}. Sanstead et al. highlighted the role of these dynamics during duplex dissociation, where the stability and relaxation timescale of frayed states was dictated by G:C base placement 10-mer oligonucleotides \citep{Sanstead2016}. Together, these dynamics represent substantial deviations from all-or-nothing behavior that can be variably expressed via subtle sequence differences in short oligomers.

In this work, we study the same four sequences explored by Sanstead et al. in an effort to uncover sequence-dependent dynamics and their relation to metastable structures mentioned above \citep{Sanstead2016}. We use the coarse-grained 3 Sites per Nucleotide (3spn2) model to simulate hybridization and dissociation behavior near each sequence's melting temperature \citep{Hinckley2013AnHybridization}. We perform these analyses without biasing simulations or assuming that one processes is a strictly reversible version of the another. Furthermore, we leverage the properties of Markov State Models (MSMs) -- namely that conditional probability depends only on the current state of the system  -- to combine many independent and unbiased trajectories and develop an understanding of sequence-specific kinetics and thermodynamics \citep{Pande2010EverythingAsk}. MSMs have recently been implemented to study mechanisms and microstate distributions of DNA hybridization, but the slowest sequence-dependent kinetics were not the focus of these studies \citep{Jin2019, Xiao2019}. \citet{Pinamonti2017} used MSMs to compare the slowest dynamics of short RNA nucleotides and found that stacking timescales are highly sequence dependent. We take a similar approach to study 10-mer DNA oligonucleotides and introduce State Free Reversible Vampnets (SRVs) to directly learn the slowest sequence-dependent dynamical modes \citep{Chen}. Furthermore, we integrate SRVs into the MSM pipeline by generating an optimized low dimensional basis in which microstate clustering can be performed. We show that SRV coordinates can be useful for both directly interpreting dynamical trends and for improving overall SRV-MSM quality when compared to more conventional methods such as time-structure independent components analysis (tICA).

We find that G:C base pair placement in decamer oligonucleotides has a substantial effect on dynamical behavior. By evaluating equilibrium trajectories we can study the relevance of metastable states during both the hybridization and dissociation process. Because SRVs generate an optimized low dimensional basis, we show that higher resolution MSMs (measured by a reduction in the required lag time) are accessible. Additionally, we compare slow dynamical modes and metastable states between sequence-specific SRV-MSMs. Within these metastable dynamical states, we leverage diffusion maps to analyze the diversity of structures whose inter-conversion rate are too fast to produce unique slow modes. Finally, we run higher temperatures simulations to investigate the temperature-dependent nature of some experimentally relevant dynamical responses. Taken together, our analysis reflects similar results to previous computational and experimental DNA work, while elucidating new insights into sequence-dependent dynamics, metastable structures, and relative timescales.

@@

The 3 site per nucleotide (3spn2) coarse grained model was designed to accelerate DNA computation relative to all-atom models while maintaining experimental melting temperatures, stacking energies, and persistence lengths\citep{Hinckley2013AnHybridization}. Interactions sites are located at phosphate, sugar, and base centers of mass; anisotropic potentials are designed to model non-bonded interactions such as intra-strand base-stacking, inter-strand cross-stacking, and base pairing. The model has been validated against experimentally determined structural properties and hybridization rates, although no dynamic information was used to parameterize the model other than Langevin friction coefficients \citep{Hinckley2013AnHybridization, Hinckley2014Coarse-grainedEffects}. 3spn2 been has widely-adopted to study numerous phenomenon including DNA packing in viral capsids, protein-DNA binding, and nucleosome unwrapping. \citep{Cordoba2017AIons, Lu2020OpenAWSEMSummary, Lequieu2016Tension-dependentUnwrapping}.

We initialized four sequences previously the subject of ultra-fast T-jump experimentation -- 5'ATATATATAT3' (AT-all), 5'GATATATATC3' (GC-end), 5'ATATGCATAT3' (GC-core), and 5'ATGATATCAT3' (GC-mix) -- along with their complementary strands in a periodic box of dimensions 7.774 nm \citep{Sanstead2016, Phys2014}. We specified a 240 mM implicit salt concentration and used a Debye-Huckel approximation for electrostatic interactions within a 5 nm cutoff. We ran our simulations in the NVT ensemble and fixed temperature via a Langevin thermostat to account for implicit solvent interactions \citep{Schneider1978Molecular-dynamicsTransitions}. Simulations were conducted at empirically determined  melting temperatures in order to maximize transitions between dissociated and hybridized states. We used a 20 fs time step and $1.3\times10^{9}$ steps in each simulation, saving frames every 100 ps. For each sequences, 40 simulations were performed in parallel, consuming about 24 serial CPU-hours on 28xIntel E5-2680v4 processors per simulation. Half of runs were initialized from the native hybridized state and half from a dissociated state. The first 10000 frames (1 $\mu$s of simulation time) of each simulation were removed, resulting in 40 x 250000 frames and a total of 1 ms simulation time per sequence. Depending on sequence, we observed between 55-100 hybridizing and dissociating transitions across all runs.

@@

All intermolecular pairwise distances for both oligonucleotides were calculated at each frame using the MDtraj software package \citep{McGibbon2015MDTraj:Trajectories}. Intermolecular distances removed translational and rotational symmetries, but additional symmetries arise from the self-complementary nature of these sequences -- both the sense and anti-sense strands are identical to each other. To account for this, we averaged permutable distances (45 pairs in total) together following a similar procedure used in tICAgg coordinates construction \citep{Sengupta2019AutomatedSelf-assembly}. The VAMP-2 scoring method was employed to evaluate the kinetic variance of the feature set and optimize hyperparameters \citep{WuVariationalData, Mardt2018VAMPnetsKinetics}. The VAMP-2 score uses the covariances of a set of inputs to estimate the transfer operator of a dynamical system, providing a robust and object means to evaluate various parameters and models. (Can include equations or additional info in SI?) We compared scores for the intermolecular, intra+intermolecular,  and symmetrized intermolecular distances, in addition to the reciprocal value for each of these coordiantes (figure \ref{fig:allseq_features_vamp2}). We found that reciprocal coordinates performed better across the board, but we saw no substantial difference between the three featurization methods. This indicated that minimal kinetic information was lost in the symmetrization procedure, therefore we chose reciprocal symmetrized coordinates to train the model. The smaller feature set enabled faster training times and better statistics over permutable distances, without appearing to suffer a loss in generality or model resolution. These features were normalized and passed into sequence-specific SRVs.

@@

@@

SRVs were first developed by Chen et al. as a means to directly learn slow eigenfunctions of the transfer operator \citep{Chen}. The method is a descendant of VAMPnets \citep{Mardt2018VAMPnetsKinetics}, deep canonical correlation analysis \citep{Andrew2013DeepAnalysis}, and extended dynamic mode decomposition with dictionary learning (EDMD-DL) \citep{Li2017ExtendedOperator}. The framework uses a twin-lobed artificial neural network to learn an optimal basis set for the variational approach to conformational dynamics (VAC) from which the leading eigenfunctions of the transfer operator are then estimated \citep{Noe2013ASystems}. The resulting orthogonal modes are associated with the slowest dynamical processes in a system, and can be used to interpret kinetic information directly (such as physical correlations and timescales) and to construct MSMs \citep{Sidky, Pande2010EverythingAsk}. SRVs provide robust nonlinear approximation and computation time that scales linearly with the amount of input data and are more powerful and efficient that tICA and ktICA approaches. This is a key attribute to our system as 10 million frames with 55 features in each frame are used for each sequence. The SRV framework has been tested on toy systems where the true eigenfunctions of the transfer operator are known and on small protein simulation data such as the WW-domain and Trp-cage mini-protein \citep{Chen, Sidky}. For these latter system, SRV-MSMs were contructed in order to find the stability of metastables states as well as transition probabilities between those states.

@@

MSMs are a powerful tool for interpreting large amounts of simulation data in a statistically robust and experimentally comparable way \citep{Phys2011MarkovValidation, Husic2018MarkovScience}. Kinetically similar conformations are first discretized into microstates and the conditional probability of transition between state is calculated within some lag time. The reliance on conditional probabilities allows for many independent simulations (longer than the lag time) to be collectively interpreted. To take full advantage of the MSM frameworks, however, the input basis should be as kinetically meaningful as possible \citep{Pande2010EverythingAsk}. Because SRV eigenfunctions translate simulation features into their slowest kinetic representations, they are optimally suited as an MSM basis. To build our SRV-MSM framework, we employed the PyEmma MSM pipeline and generated independent models for each sequence \citep{Scherer2015PyEMMAModels}. In a similar approach to Sidky et al. we performed k-means microstate clustering, Bayesian MSM construction, and PCCA+ hierarchical macrostate assignments \citep{Sidky}. The number of microstates were determined by VAMP-2 score, and the SRV-MSM lag time was selected based on implied timescales convergence.  The number of PCCA+ macrostates was determined based on the characteristic of each system and will be discussed more in depth in the results and supplemental information.

@@

@@

\subsection{Interpreting PCCA+ Macrostates}

@@

\subsection{SRV training}

\subsection{SRV-MSM walkthrough and analysis}
In our analysis, we found that the AT-all sequence, given its repetitive structure and lack of GC-content, produced the simplest dynamics and displayed a clear spectral gap between modes. For this reason, we use this sequence a case study to work through our SRV-MSM pipeline step-by-step. Our first task was to identify the SRV lag time that was longer than the intrinsic Markov timescales of the system, yet short enough to resolve the dynamics of interest \citep{Phys2018MarkovValidation}. We found that most implied timescales converge at an SRV lag time of 1.2 ns. Next we selected an optimal number of SRV components to include in our analysis. After a certain point, higher order dynamical modes provide diminishing contributions the overall kinetic variance as measured by the Vamp-2 score, and the model can begin fitting on statistical noise in the trajectory data instead of the true dynamics \citep{McGibbon2015VariationalKinetics}. It is also more difficult to perform kmeans clustering on a high dimensional space, especially when those higher dimensions are less kinetically relevant \citep{Pande2010EverythingAsk}. For these reasons, the number of slow SRV components should be carefully selected based on the specific system of interest. As shown in figure \ref{fig:allseq_srv_crossval}), we see diminishing returns in the VAMP-2 score after five slow modes and select these modes as our optimized SRV basis.
